---
title: "Dissertation Data Analysis"
author: "Maddy Tattersall"
format: 
 html:
   toc: true
editor: visual
execute: 
  warning: false
  message: false
---

## Load packages

```{r}
library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(cowplot)
library(ggfortify)
library(lme4)
library(rstatix)
library(ggplot2)
library(cowplot)
```

## Load data

```{r}
filepath_data <- here("Data collection","Results.xlsx")
mydata <- read_excel(filepath_data,"Sheet2")
glimpse(mydata)
```

## pH Analysis

### Summarise

```{r}
pH_summary <- mydata |>
  group_by(Position) |>
  summarise(n = n(), mean_pH = mean(pH), sd_pH = sd(pH), se_pH = sd(pH)/sqrt(n()))
pH_summary
```

### Plot

**Bar Chart**

```{r}
pH_summary |>
  ggplot(aes(x = Position, y = mean_pH)) +
  geom_col() + 
  geom_errorbar(aes(ymin = mean_pH - se_pH, ymax= mean_pH + se_pH), width = 0.15) +
  labs(x = "",
       y = "mean pH") +
  theme_cowplot()
```

**Box Plot**

```{r}
ggplot(mydata, aes(x = Position, y = pH)) +
  geom_boxplot() +
  geom_jitter(width = 0.1) +
  labs(
    x = "",
    y = "pH",
    title = "pH Above and Below Beaver Dam"
  ) +
  cowplot::theme_cowplot() +
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15)
  )
```

**Violin Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y = pH, fill = Position)) +
  geom_violin(
    trim = FALSE,
    alpha = 0.6
  ) +
  scale_fill_manual(
    values = c(
      "Below" = "#00BFC4",  # same blue as line graphs
      "Above" = "red"
    )
  ) +
  labs(
    x = "",
    y = "pH",
    title = "pH Above and Below Beaver Dam"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "none",  # 👈 removes legend
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    )
  )
```

**Violin with Box Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y = pH)) +
  
  geom_violin(
    aes(fill = Position),
    trim = FALSE,
    alpha = 0.6
  ) +
  
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    fill = "white",
    colour = "black"
  ) +
  
  scale_fill_manual(
    values = c(
      "Above" = "red",
      "Below" = "#00BFC4"
    )
  ) +
  
  labs(
    x = "",
    y = "pH",
    title = "pH Above and Below Beaver Dam"
  ) +
  
  theme_cowplot() +
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    ),
    legend.position = "none"
  )
```

### Analysis

**Shapiro-Wilk Test**

```{r}
mydata |>
  group_by(Position) |>
  shapiro_test(pH)
```

**Wilcox-test**

```{r}
mydata |>
  wilcox_test(pH~Position)
```

**Mixed effects analysis**

```{r}
me_pH <- lmer(pH ~ Position + (1|Month), data = mydata)
summary(me_pH)
```

## Temperature Analysis

### Summarise

```{r}
Temperature_summary <- mydata |>
  group_by(Position) |>
  summarise(n = n(), mean_Temperature = mean(Temperature), sd_Temperature = sd(Temperature), se_Temperature = sd(Temperature)/sqrt(n()))
Temperature_summary
```

### Plot

**Bar Chart**

```{r}
Temperature_summary |>
  ggplot(aes(x = Position, y = mean_Temperature)) +
  geom_col() + 
  geom_errorbar(aes(ymin = mean_Temperature - se_Temperature, ymax= mean_Temperature + se_Temperature), width = 0.15) +
  labs(x = "",
       y = "mean Temperature") +
  theme_cowplot()
```

**Box Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y =Temperature)) +
  geom_boxplot() + 
  geom_jitter() +
  labs(x = "",
       y = "Temperature (ºC)",
       title="Temperature Above and Below Beaver Dam") +
  theme_cowplot() +
  theme (
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      size = 15, 
      hjust = 0.5,
      face = "bold"
    )
  )
  
```

**Violin Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y = Temperature, fill = Position)) +
  geom_violin(
    trim = FALSE,
    alpha = 0.6
  ) +
  scale_fill_manual(
    values = c(
      "Below" = "#00BFC4",  # same blue as line graphs
      "Above" = "red"
    )
  ) +
  labs(
    x = "",
    y = "Temperature (ºC)",
    title = "Temperature Above and Below Beaver Dam"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "none",  # 👈 removes legend
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    )
  )
```

**Violin with Box Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y = Temperature)) +
  
  geom_violin(
    aes(fill = Position),
    trim = FALSE,
    alpha = 0.6
  ) +
  
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    fill = "white",
    colour = "black"
  ) +
  
  scale_fill_manual(
    values = c(
      "Above" = "red",
      "Below" = "#00BFC4"
    )
  ) +
  
  labs(
    x = "",
    y = "Temperature (ºC)",
    title = "Temperature Above and Below Beaver Dam"
  ) +
  
  theme_cowplot() +
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    ),
    legend.position = "none"
  )
```

### Analysis

**Shapiro-Wilk Test**

```{r}
mydata |>
  group_by(Position) |>
  shapiro_test(Temperature)
```

**Wilcox-test**

```{r}
mydata |>
  wilcox_test(Temperature~Position)
```

## TDS Analysis

### Summarise

```{r}
TDS_summary <- mydata |>
  group_by(Position) |>
  summarise(n = n(), mean_TDS= mean(TDS), sd_TDS = sd(TDS), se_TDS = sd(TDS)/sqrt(n()))
TDS_summary
```

### Plot

**Bar Chart**

```{r}
TDS_summary |>
  ggplot(aes(x = Position, y = mean_TDS)) +
  geom_col() + 
  geom_errorbar(aes(ymin = mean_TDS - se_TDS, ymax= mean_TDS + se_TDS), width = 0.15) +
  labs(x = "",
       y = "mean TDS") +
  theme_cowplot()
```

**Box Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y =TDS)) +
  geom_boxplot() + 
  geom_jitter() +
  labs(x = "",
       y = "TDS (ppm)", 
  title ="TDS Above and Below Beaver Dam") +
  theme_cowplot() +
  theme (
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      size = 15, 
      hjust = 0.5,
      face = "bold"
    )
  )
```

**Violin Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y = TDS, fill = Position)) +
  geom_violin(
    trim = FALSE,
    alpha = 0.6
  ) +
  scale_fill_manual(
    values = c(
      "Below" = "#00BFC4",  # same blue as line graphs
      "Above" = "red"
    )
  ) +
  labs(
    x = "",
    y = "TDS (ppm)",
    title = "TDS Above and Below Beaver Dam"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "none",  # 👈 removes legend
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    )
  )
```

**Violin with Box Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y = TDS)) +
  
  geom_violin(
    aes(fill = Position),
    trim = FALSE,
    alpha = 0.6
  ) +
  
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    fill = "white",
    colour = "black"
  ) +
  
  scale_fill_manual(
    values = c(
      "Above" = "red",
      "Below" = "#00BFC4"
    )
  ) +
  
  labs(
    x = "",
    y = "TDS (ppm) ",
    title = "TDS Above and Below Beaver Dam"
  ) +
  
  theme_cowplot() +
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    ),
    legend.position = "none"
  )
```

### Analysis

**Shapiro-Wilk Test**

```{r}
mydata |>
  group_by(Position) |>
  shapiro_test(TDS)
```

**t-test**

```{r}
mydata |>
  t_test(TDS~Position)
```

## TON Analysis

### Summarise

```{r}
Nitrates_summary <- mydata |>
  group_by(Position) |>
  summarise(n = n(), mean_Nitrates= mean(Nitrates), sd_Nitrates = sd(Nitrates), se_Nitrates = sd(Nitrates)/sqrt(n()))
Nitrates_summary
```

### Plot

**Bar Chart**

```{r}
Nitrates_summary |>
  ggplot(aes(x = Position, y = mean_Nitrates)) +
  geom_col() + 
  geom_errorbar(aes(ymin = mean_Nitrates - se_Nitrates, ymax= mean_Nitrates + se_Nitrates), width = 0.15) +
  labs(x = "",
       y = "mean Nitrates") +
  theme_cowplot()
```

**Box Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y =Nitrates)) +
  geom_boxplot() + 
  geom_jitter() +
  labs(x = "",
            y = "TON (mg/l)", 
  title ="TON Above and Below Beaver Dam") +
  theme_cowplot() +
  theme (
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      size = 15, 
      hjust = 0.5,
      face = "bold",
    )
  )
```

**Violin Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y = Nitrates, fill = Position)) +
  geom_violin(
    trim = FALSE,
    alpha = 0.6
  ) +
  scale_fill_manual(
    values = c(
      "Below" = "#00BFC4",  # same blue as line graphs
      "Above" = "red"
    )
  ) +
  labs(
    x = "",
    y = "TON (mg/l)",
    title = "TON Above and Below Beaver Dam"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "none",  # 👈 removes legend
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    )
  )
```

**Violin with Box Plot**

```{R}
mydata |>
  ggplot(aes(x = Position, y = Nitrates)) +
  
  geom_violin(
    aes(fill = Position),
    trim = FALSE,
    alpha = 0.6
  ) +
  
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    fill = "white",
    colour = "black"
  ) +
  
  scale_fill_manual(
    values = c(
      "Above" = "red",
      "Below" = "#00BFC4"
    )
  ) +
  
  labs(
    x = "",
    y = "TON mg/l",
    title = "TON Above and Below Beaver Dam"
  ) +
  
  theme_cowplot() +
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    ),
    legend.position = "none"
  )
```

### Analysis

**Shapiro-Wilk Test**

```{r}
mydata |>
  group_by(Position) |>
  shapiro_test(Nitrates)
```

**Wilcox-test**

```{r}
mydata |>
  wilcox_test(Nitrates~Position)
```

## DO Analysis

### Summarise

```{r}
DO_summary <- mydata |>
  group_by(Position) |>
  summarise(n = n(), mean_DO= mean(DO), sd_DO = sd(DO), se_DO = sd(DO)/sqrt(n()))
DO_summary
```

### Plot

**Bar Chart**

```{r}
DO_summary |>
  ggplot(aes(x = Position, y = mean_DO)) +
  geom_col() + 
  geom_errorbar(aes(ymin = mean_DO - se_DO, ymax= mean_DO + se_DO), width = 0.15) +
  labs(x = "",
       y = "mean DO") +
  theme_cowplot()
```

**Box Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y =DO)) +
  geom_boxplot() + 
  geom_jitter() +
  labs(x = "",
        y = "Dissolved Oxygen (%)", 
  title ="Dissolved Oxygen Above and Below Beaver Dam") +
  theme_cowplot() +
  theme (
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      size = 15, 
      hjust = 0.5,
      face = "bold"
    )
  )
```

**Violin Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y = DO, fill = Position)) +
  geom_violin(
    trim = FALSE,
    alpha = 0.6
  ) +
  scale_fill_manual(
    values = c(
      "Below" = "#00BFC4",  # same blue as line graphs
      "Above" = "red"
    )
  ) +
  labs(
    x = "",
    y = "DO (%)",
    title = "DO Above and Below Beaver Dam"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "none",  # 👈 removes legend
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    )
  )
```

**Violin with Box Plot**

```{r}
mydata |>
  ggplot(aes(x = Position, y = DO)) +
  
  geom_violin(
    aes(fill = Position),
    trim = FALSE,
    alpha = 0.6
  ) +
  
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    fill = "white",
    colour = "black"
  ) +
  
  scale_fill_manual(
    values = c(
      "Above" = "red",
      "Below" = "#00BFC4"
    )
  ) +
  
  labs(
    x = "",
    y = "DO (%)",
    title = "DO Above and Below Beaver Dam"
  ) +
  
  theme_cowplot() +
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 15
    ),
    legend.position = "none"
  )
```

### Analysis

**Shapiro-Wilk Test**

```{r}
mydata |>
  group_by(Position) |>
  shapiro_test(DO)
```

**Wilcox-test**

```{r}
mydata |>
  wilcox_test(DO~Position)
```

### Checking for Normality

**Change parameter accordingly**

```{r}
hist(mydata$pH, 
     main = "Histogram of pH", 
     xlab = "pH", 
     col = "lightblue", 
     border = "black")
```

## Line graphs

**Ensure month order**

```{r}
mydata$Month <- factor(mydata$Month, 
                   levels = c("April", "May", "June", "July", "August", "September"))
```

**Summarise**

```{r}
summary_mydata <- mydata %>%
  group_by(Month, Position) %>%
  summarise(
    pH = mean(pH, na.rm = TRUE),
    Temperature = mean(Temperature, na.rm = TRUE),
    TDS = mean(TDS, na.rm = TRUE),
    DO = mean(DO, na.rm = TRUE),
    Nitrates = mean(Nitrates, na.rm = TRUE)
  ) %>%
  ungroup()
```

**pH**

```{r}
ggplot(summary_mydata, aes(x = Month, y = pH, color = Position, group = Position)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Average pH Over Months", y = "pH") +
  theme_minimal()+
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  )
```

**TDS**

```{r}
ggplot(summary_mydata, aes(x = Month, y = Temperature, color = Position, group = Position)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Average Temperature Over Months", y = "Temperature (ºC)") +
  theme_minimal()+
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  )
```

**DO**

```{r}
ggplot(summary_mydata, aes(x = Month, y = DO, color = Position, group = Position)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Average DO Over Months", y = "DO (%)") +
  theme_minimal()+
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  )
```

**TDS**

```{r}
ggplot(summary_mydata, aes(x = Month, y = TDS, color = Position, group = Position)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Average TDS Over Months", y = "TDS (ppm)") +
  theme_minimal()+
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  )
```

**Nitrates**

```{r}
ggplot(summary_mydata, aes(x = Month, y = Nitrates, color = Position, group = Position)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Average TON Over Months", y = "TON (mg/l)") +
  theme_minimal()+
  theme(
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  )
```
